---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# drugdevelopR: Utility-based optimal phase II/III drug development planning.

<!-- badges: start -->
[![Travis build status](https://travis-ci.com/Sterniii3/drugdevelopR.svg?branch=master)](https://travis-ci.com/Sterniii3/drugdevelopR)
[![CRAN status](https://www.r-pkg.org/badges/version/drugdevelopR)](https://CRAN.R-project.org/package=drugdevelopR)
[![R-CMD-check](https://github.com/Sterniii3/drugdevelopR/workflows/R-CMD-check/badge.svg)](https://github.com/Sterniii3/drugdevelopR/actions)
[![Codecov test coverage](https://codecov.io/gh/Sterniii3/drugdevelopR/branch/master/graph/badge.svg)](https://app.codecov.io/gh/Sterniii3/drugdevelopR?branch=master)
<!-- badges: end -->

The drugdevelopR package enables you to plan phase II/III drug development programs with optimal sample size allocation and go/no-go decision rules. The assumed true treatment effects can be fixed or modelled by a prior distribution. The corresponding [R Shiny application](https://web.imbi.uni-heidelberg.de/drugdevelopR/) has a graphic user interface for the package and thus makes it accessible for users without prior knowledge of R. Fast computing is made possible by parallel programming. The theoretical foundations for this package were laid in the dissertation  "Integrated Planning of Pilot and Subsequent Confirmatory Study in Clinical Research – Finding Optimal Designs in a Utility-Based Framework"  by Stella Erdmann at the Institute of Medical Biometry at the University of Heidelberg.

On the package webpage, we supply [full documentation](https://sterniii3.github.io/drugdevelopR/reference/index.html) of all functions as well as a [tutorial for getting started](https://sterniii3.github.io/drugdevelopR/vignettes/introduction-to-drugdevelopr.html) with drugdevelopR.

## Installation

Install the development version of the package directly from [GitHub](https://github.com/Sterniii3/drugdevelopR/) using the following code:
```r
if(!require(devtools)) { install.packages("devtools"); require(devtools)} 
devtools::install_github("Sterniii3/drugdevelopR")
````
and access the drugdevelopR App via [https://web.imbi.uni-heidelberg.de/drugdevelopR/](https://web.imbi.uni-heidelberg.de/drugdevelopR/).

## Usage

Here is a basic example for applying drugdevelopR to a drug development program with a normally distributed outcome:
```{r}
library(drugdevelopR)
# TODO: Fill example here.
```
