---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# drugdevelopR: Utility-based optimal phase II/III drug development planning.

<!-- badges: start -->
[![Travis build status](https://travis-ci.com/Sterniii3/drugdevelopR.svg?branch=master)](https://travis-ci.com/Sterniii3/drugdevelopR)
[![CRAN status](https://www.r-pkg.org/badges/version/drugdevelopR)](https://CRAN.R-project.org/package=drugdevelopR)
[![R-CMD-check](https://github.com/Sterniii3/drugdevelopR/workflows/R-CMD-check/badge.svg)](https://github.com/Sterniii3/drugdevelopR/actions)
[![Codecov test coverage](https://codecov.io/gh/Sterniii3/drugdevelopR/branch/master/graph/badge.svg)](https://app.codecov.io/gh/Sterniii3/drugdevelopR?branch=master)
<!-- badges: end -->

The drugdevelopR package enables you to plan phase II/III drug development programs with optimal sample size allocation and go/no-go decision rules. The assumed true treatment effects can be fixed or modelled by a prior distribution. The corresponding [R Shiny application](https://web.imbi.uni-heidelberg.de/drugdevelopR/) has a graphic user interface for the package and thus makes it accessible for users without prior knowledge of R. Fast computing is made possible by parallel programming. The theoretical foundations for this package were laid in the dissertation  "Integrated Planning of Pilot and Subsequent Confirmatory Study in Clinical Research – Finding Optimal Designs in a Utility-Based Framework"  by Stella Erdmann at the Institute of Medical Biometry at the University of Heidelberg.

On the package webpage, we supply [full documentation](https://sterniii3.github.io/drugdevelopR/reference/index.html) of all functions as well as a [tutorial for getting started](https://sterniii3.github.io/drugdevelopR/vignettes/introduction-to-drugdevelopr.html) with drugdevelopR.

## Installation

Install the development version of the package directly from [GitHub](https://github.com/Sterniii3/drugdevelopR/) using the following code:
```r
if(!require(devtools)) { install.packages("devtools"); require(devtools)} 
devtools::install_github("Sterniii3/drugdevelopR")
````
and access the drugdevelopR App via [https://web.imbi.uni-heidelberg.de/drugdevelopR/](https://web.imbi.uni-heidelberg.de/drugdevelopR/).

## Usage

Here is a basic example for applying drugdevelopR to a drug development program with a normally distributed outcome:
```{r}
library(drugdevelopR)
res <- optimal_normal(Delta1 = 0.625, Delta2 = 0.8, fixed = FALSE, # treatment effect
                      n2min = 20, n2max = 400, # sample size region
                      stepn2 = 4, # sample size step size
                      kappamin = 0.02, kappamax = 0.2, # threshold region
                      stepkappa = 0.02, # threshold step size
                      c2 = 0.675, c3 = 0.72, # maximal total trial costs
                      c02 = 15, c03 = 20, # maximal per-patient costs
                      b1 = 3000, b2 = 8000, b3 = 10000, # gains for patients
                      alpha = 0.05, # significance level
                      beta = 0.1, # 1 - power
                      w = 0.6, in1 = 300, in2 = 600, # weight and amount of information
                      a = 0.25, b = 0.75) # truncation values
```

## drugdevelopR functions
The drugdevelopR package provides the functions

* [`optimal_tte`](https://sterniii3.github.io/drugdevelopR/reference/optimal_tte.html),
* [`optimal_binary`](https://sterniii3.github.io/drugdevelopR/reference/optimal_binary.html), and
* [`optimal_normal`](https://sterniii3.github.io/drugdevelopR/reference/optimal_normal.html)

to plan optimal phase II/III drug development programs with

* time-to-event (treatment effect measured by hazard ratio (HR)),
* binary (treatment effect measured by risk ratio (RR)), or
* normally distributed (treatment effect measured by standardized difference in means (Delta))

endpoints, where the treatment effect is modelled by a [prior](https://web.imbi.uni-heidelberg.de/prior/). Optimal phase II/III drug development planning with fixed treatment effects can be done with the help of the R Shiny application [basic](https://web.imbi.uni-heidelberg.de/basic/).

Extensions to the basic setting are:

* optimal planning of programs including methods for discounting of phase II results (function: [optimal_bias](https://sterniii3.github.io/drugdevelopR/reference/optimal_bias.html), App: [bias](https://web.imbi.uni-heidelberg.de/bias/)),
* optimal planning of programs with several phase III trials (function: [optimal_multitrial](https://sterniii3.github.io/drugdevelopR/reference/optimal_multitrial.html), App: [multitrial](https://web.imbi.uni-heidelberg.de/multitrial/)) and
* optimal planning of programs with multiple arms (function: [optimal_multiarm](https://sterniii3.github.io/drugdevelopR/reference/optimal_multiarm.html), App: [multiarm](https://web.imbi.uni-heidelberg.de/multiarm/)).
