---
title: "More Parameters"
author: "Johannes Cepicka"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{More_Parameters}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Suppose we are planning a drug development program testing the superiority of an experimental treatment over a control treatment. Our drug development program consists of an exploratory phase II trial which is, in case of promising results, followed by a confirmatory phase III trial.

The drugdevelopR package enables us to optimally plan such programs using a utility-maximizing approach. To get a brief introduction, we presented a very basic example on how the package works in [Introduction to planning phase II and phase III trials with drugdevelopR](https://sterniii3.github.io/drugdevelopR/articles/introduction-to-drugdevelopR.html). However, there are many more parameters, which the user can provide, to adapt the program to your specific needs. In this article, we therefore want to present the options you have.

# The example setting and the input parameters

We are in the same setting as in the introduction, i.e. we  suppose we are developing a new tumor treatment, *exper*. The patient variable that we want to investigate is the difference in tumor width between the one-year visit and baseline. This is a normally distributed outcome variable.

The parameters we insert into the function `optimum_normal` are the same parameters we also inserted in the basic setting. Thus, for more information on the input values we refer to the [Introduction](https://sterniii3.github.io/drugdevelopR/articles/introduction-to-drugdevelopR.html).
We start by loading the drugdevelopR package.
```{r setup}
library(drugdevelopR)
```

Back then, we got the following output:

```{r, eval=TRUE, include=FALSE}
res <- readRDS(file="optimal_normal_basic_setting.RDS")
```
```{r}
res
```

We now want to see, how different modifications, i.e using more paramters change the outcome of the program. Note, that setting the parameter `fixed` to be `"False"`, and thus using a prior distribution, was already explained in the article 
[*Fixed or prior*](https://sterniii3.github.io/drugdevelopR/articles/Fixed_and_prior_distributions.html)

## Constraint Optimization
The are three options how to include constraint optimization, by setting maximum costs limits, a maximum number of participants or by demanding a minimum probability of a successful program. These options can be applied separately or combined, as it is best suited to your specific needs.

### Cost Constraint - Maximum Total Costs
The total costs are given as sum of the cost ins phase II and phase III, i.e. as $K2$ + $K3$. In our example we get total costs of $72+139=211$ (in 10^5 \$). Let's assume an investor in your drug or a pharmaceutical company is only willing to pay 20,000,000 \$. You can now set a maximum cost constraint by providing an parameter value of `K = 200` into the `optimum_normal` function. The default value for `K` is `Inf`.

```{r,eval = FALSE}
 resK <- optimal_normal(Delta1 = 0.625, fixed = TRUE, # treatment effect
                       n2min = 20, n2max = 400, # sample size region
                       stepn2 = 4, # sample size step size
                       kappamin = 0.02, kappamax = 0.2, # threshold region
                       stepkappa = 0.02, # threshold step size
                       c2 = 0.675, c3 = 0.72, # maximal total trial costs
                       c02 = 15, c03 = 20, # maximal per-patient costs
                       b1 = 3000, b2 = 8000, b3 = 10000, # gains for patients
                       alpha = 0.05, # significance level
                       beta = 0.1, # 1 - power
                       Delta2 = NULL, w = NULL, in1 = NULL, in2 = NULL, 
                       a = NULL,b = NULL, # setting all unneeded parameters to NULL
                       K = 200 # cost constraint
                       ) 
```

We get the following results:
```{r, eval=TRUE, include=FALSE}
# Comment this chunk after running it once
#  resK <- optimal_normal(Delta1 = 0.625, fixed = TRUE, # treatment effect
#                        n2min = 20, n2max = 400, # sample size region
#                        stepn2 = 4, # sample size step size
#                        kappamin = 0.02, kappamax = 0.2, # threshold region
#                        stepkappa = 0.02, # threshold step size
#                        c2 = 0.675, c3 = 0.72, # maximal total trial costs
#                        c02 = 15, c03 = 20, # maximal per-patient costs
#                        b1 = 3000, b2 = 8000, b3 = 10000, # gains for patients
#                        alpha = 0.05, # significance level
#                        beta = 0.1, # 1 - power
#                        Delta2 = NULL, w = NULL, in1 = NULL, in2 = NULL, 
#                        a = NULL,b = NULL,
#                        K = 200) # setting all unneeded parameters to NULL
# saveRDS(resK, file="optimal_normal_cost_constraint.RDS")
```
```{r, eval=TRUE, include=FALSE}
resK <- readRDS(file="optimal_normal_cost_constraint.RDS")
```
```{r}
resK
```
We see that due to the cost constraint the costs in phase II actually increase from 72 to 74, however the costs in phase III decrease from 139 to 125, so overall the constraint is met. Moreover, the expected utility at the optimum decreases from 325,983,000 to 324,742,000 dollars.

### Sample Size Constraint - Maximum Number of Participants
Very similarly to to the cost constraint, we can set a constraint on the maximum number of participants in our trial. As we can see, without constraints, in the optimum we need a total number of 250 participants. However, one can certainly imagine a situation, where the only 200 persons willing to participate in a trial, for example because the incidence of the illness is not very high within the target population. To model this, we can easily set a sample size constraint, in this case of `N = 200`. This allows for an optimal allocation of the participants between phase II and III. The default value for `N` is `Inf`.
```{r,eval = FALSE}
 resN <- optimal_normal(Delta1 = 0.625, fixed = TRUE, # treatment effect
                       n2min = 20, n2max = 400, # sample size region
                       stepn2 = 4, # sample size step size
                       kappamin = 0.02, kappamax = 0.2, # threshold region
                       stepkappa = 0.02, # threshold step size
                       c2 = 0.675, c3 = 0.72, # maximal total trial costs
                       c02 = 15, c03 = 20, # maximal per-patient costs
                       b1 = 3000, b2 = 8000, b3 = 10000, # gains for patients
                       alpha = 0.05, # significance level
                       beta = 0.1, # 1 - power
                       Delta2 = NULL, w = NULL, in1 = NULL, in2 = NULL, 
                       a = NULL,b = NULL, # setting all unneeded parameters to NULL
                       N = 200 # sample size constraint
                       ) 
```

We get the following results:
```{r, eval=TRUE, include=FALSE}
# Comment this chunk after running it once
#  resN <- optimal_normal(Delta1 = 0.625, fixed = TRUE, # treatment effect
#                        n2min = 20, n2max = 400, # sample size region
#                        stepn2 = 4, # sample size step size
#                        kappamin = 0.02, kappamax = 0.2, # threshold region
#                        stepkappa = 0.02, # threshold step size
#                        c2 = 0.675, c3 = 0.72, # maximal total trial costs
#                        c02 = 15, c03 = 20, # maximal per-patient costs
#                        b1 = 3000, b2 = 8000, b3 = 10000, # gains for patients
#                        alpha = 0.05, # significance level
#                        beta = 0.1, # 1 - power
#                        Delta2 = NULL, w = NULL, in1 = NULL, in2 = NULL, 
#                        a = NULL,b = NULL,
#                        N = 200) 
# saveRDS(resN, file="optimal_normal_sample_size_constraint.RDS")
```
```{r, eval=TRUE, include=FALSE}
resN <- readRDS(file="optimal_normal_sample_size_constraint.RDS")
```
```{r}
resN
```
Due to our sample size constraint to optimal number of participants in phase II is 72 and 128 in phase III (down from 84 and 166 respectively) within this framework. The expected utility in the optimum decreases to 314,241,000.

### Success Probability Constraint - Minimum Probability of a Successful Program
The last option provided for constraint optimization is the minimum probability of a successful program. In our basic setting, without constraints, we hat a probability that the program was successful of 0.85, i.e. we had a probability of 0.85 that the program would proceed from phase II to phase III and that a positive treatment effect could be detected in phase III. Let's say, you are willing to have higher costs and in return want this probability to be 0.87. You can include this within the framework, by including the parameter `S` and by setting it to `S = 0.87`. The default value for `S` is `-Inf`.

```{r,eval = FALSE}
 resS <- optimal_normal(Delta1 = 0.625, fixed = TRUE, # treatment effect
                       n2min = 20, n2max = 400, # sample size region
                       stepn2 = 4, # sample size step size
                       kappamin = 0.02, kappamax = 0.2, # threshold region
                       stepkappa = 0.02, # threshold step size
                       c2 = 0.675, c3 = 0.72, # maximal total trial costs
                       c02 = 15, c03 = 20, # maximal per-patient costs
                       b1 = 3000, b2 = 8000, b3 = 10000, # gains for patients
                       alpha = 0.05, # significance level
                       beta = 0.1, # 1 - power
                       Delta2 = NULL, w = NULL, in1 = NULL, in2 = NULL, 
                       a = NULL,b = NULL, # setting all unneeded parameters to NULL
                       S = 0.87 #minimum success probability
                       ) 
```
We get the following results:
```{r, eval=TRUE, include=FALSE}
# Comment this chunk after running it once
#  resS <- optimal_normal(Delta1 = 0.625, fixed = TRUE, # treatment effect
#                        n2min = 20, n2max = 400, # sample size region
#                        stepn2 = 4, # sample size step size
#                        kappamin = 0.02, kappamax = 0.2, # threshold region
#                        stepkappa = 0.02, # threshold step size
#                        c2 = 0.675, c3 = 0.72, # maximal total trial costs
#                        c02 = 15, c03 = 20, # maximal per-patient costs
#                        b1 = 3000, b2 = 8000, b3 = 10000, # gains for patients
#                        alpha = 0.05, # significance level
#                        beta = 0.1, # 1 - power
#                        Delta2 = NULL, w = NULL, in1 = NULL, in2 = NULL, 
#                        a = NULL,b = NULL,
#                        S = 0.87) # setting all unneeded parameters to NULL
# saveRDS(resS, file="optimal_normal_probability_constraint.RDS")
```
```{r, eval=TRUE, include=FALSE}
resS <- readRDS(file="optimal_normal_probability_constraint.RDS")
```
```{r}
resS
```
The probability of a successful program now increased from 0.85 to 0.87, mainly by increasing the number of participants in phase II. The expected utility decreased to 323,557,000.

## Customized Effect Size Categories
Up until now, the effect size categories for a small, medium or large treatment effect were predefined and set to $0-0.5$ for a small treatment effect, $0.5-0.8$ for a medium treatment effect and $\geq 0.8$ for a large treatment effect (each in standardized differences in mean). These parameter values were taken from Cohen (1988). However, one can easily customize the effect size categories by inserting threshold values for each effect size. Let's say you only consider a program successful if the treatment effect is above 0.1. Furthermore, you want to assign a medium benefit if the treatment effect is between 0.6 and 1 and a large benefit if the treatment effect is above 1. Hereby, note that there is a one-to-one relation between effect size categories and benefit categories, i.e. a large treatment effect corresponds to a large benefit and so on. If you want to implement this scenario you can do this by defining the parameters `steps1`, `stepm1` and `stepl1`. Here, we set them to `steps1 = 0.1`, `stepm1 = 0.6` and `stepl1 = 1`.

```{r,eval = FALSE}
 res <- optimal_normal(Delta1 = 0.625, fixed = TRUE, # treatment effect
                       n2min = 20, n2max = 400, # sample size region
                       stepn2 = 4, # sample size step size
                       kappamin = 0.02, kappamax = 0.2, # threshold region
                       stepkappa = 0.02, # threshold step size
                       c2 = 0.675, c3 = 0.72, # maximal total trial costs
                       c02 = 15, c03 = 20, # maximal per-patient costs
                       b1 = 3000, b2 = 8000, b3 = 10000, # gains for patients
                       alpha = 0.05, # significance level
                       beta = 0.1, # 1 - power
                       Delta2 = NULL, w = NULL, in1 = NULL, in2 = NULL, 
                       a = NULL,b = NULL, # setting all unneeded parameters to NULL
                       steps1 = 0.1, stepm1 = 0.6, stepl1 = 1 #stepsizes for effect size categories
                       ) 
```

## Differences in the Treatments Effect
Due to different population structures in phase II and phase III the assumed true treatment effect `Delta1` may not be the same in both phases. Trials are often conducted in populations that are more heterogeneous than that of the pre- ceding phase II trials (Kirby et al., 2012), so the effect in phase III may be different than in phase II. In order to model for these differences, one can use the parameter `gamma`. The treatment effect in phase III is than calculated as follows: $\Delta_3 = \Delta_2 + \gamma$ Values of gamma below zero indicate a more pessimistic view, values above zero indicate a more optimistic view about the treatment effect.

## Skipping Phase II

## Parallel Computing

# Where to go from here
This tutorial presents even more parameters, so you can adapt the the program to your specific needs.
For more information on how to use the package, see:

* [*Introduction to drugdevelopR:*](https://sterniii3.github.io/drugdevelopR/articles/introduction-to-drugdevelopR.html) See how the package works in a basic example.
* *Different outcomes:* Apply it to binary endpoints and time-to-event endpoints.
* [*Interpreting the rest of the output:*](https://sterniii3.github.io/drugdevelopR/articles/Interpreting_Output.html) Obtain further details on your drug development program.
* [*Fixed or prior:*](https://sterniii3.github.io/drugdevelopR/articles/Fixed_and_prior_distributions.html) Model the assumed treatment effect on a prior distribution.
* *Complex drug development programs:* Adapt to situations with biased effect estimators, multiple phase III trials, multi-arm trials, or multiple endpoints.

# References
Cohen, J. (1988). Statistical power analysis for the behavioral sciences.

Kirby, S., Burke, J., Chuang-Stein, C., and Sin, C. (2012). Discounting phase 2 results when planning phase 3 clinical trials. Pharmaceutical Statistics, 11(5):373–385.
