---
title: "Programs for Trials with Multiple Endpoints"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multiple_Endpoints}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Suppose we are planning a drug development program testing the superiority of an experimental treatment over a control treatment. Our drug development program consists of an exploratory phase II trial which is, in case of promising results, followed by a confirmatory phase III trial.

The drugdevelopR package enables us to optimally plan such programs using a utility-maximizing approach. To get a brief introduction, we presented a very basic example on how the package works in [Introduction to planning phase II and phase III trials with drugdevelopR](https://sterniii3.github.io/drugdevelopR/articles/introduction-to-drugdevelopR.html). Contrary to the basic setting, when only one phase II and one phase III trial were conducted, in this scenario we investigate what happens when multiple endpoints are of interest. The functions currently implemented allow for the conduction of trials with two different endpoints. The drugdevelopR package enables programs where only one endpoints has to show a significant positive treatment effect or all endpoints have to show a significant positive treatment effect.  As the differences between these two cases are more pronounced than in the other methods implemented for the package, we present examples for both settings individually in this article.

# Only one endpoint has to be significant

Suppose we are developing a new tumor treatment, *exper*.  The two endpoints we consider interesting are overall survival (OS) and progression free survival (PFS). These are time-to-event outcome variables. Therefore, we will use the function `optimal_multiple_tte`, which calculates optimal sample sizes and threshold decisions values for time-to-event outcomes when multiple endpoints are investigated. 

Within our drug development program, we will compare our experimental treatment to the control treatment *contro*. The treatment effect measure of *exper* for each endpoint $i \in \{1,2\}$ is given by $\theta_i = −\log(HR_i)$, which is the negative logarithm of the hazard ratio $HR_i$, which is the ratio of the hazard rates between the experimental treatment and the control treatment, for each endpoint. For more information on the hazard ratio, see the vignette on [time-to-event endpoints](https://sterniii3.github.io/drugdevelopR/articles/Time-to-event_outcomes.html). 

## Applying the package to the example

After installing the package according to the [installation instructions](https://sterniii3.github.io/drugdevelopR/#Installation), we can load it using the following code:
```{r}
library(drugdevelopR)
```

## Defining all necessary parameters
As, the parameters in the setting with multiple endpoints differ slightly from the other cases ([bias adjustment](https://sterniii3.github.io/drugdevelopR/articles/Bias_adjustment.html) and [multitrial](https://sterniii3.github.io/drugdevelopR/articles/Multitrial.html)), we will explain the differing parameters in more detail.
As in the basic setting, the treatment effect may be fixed (as in this example) or follows a prior distribution (see [Fixed or Prior](https://sterniii3.github.io/drugdevelopR/articles/Fixed_and_prior_distributions.html)) Note, that in this case, contrary to the other setting, the second treatment effect describes the effect of the second endpoint and is also relevant for `fixed = TRUE`. Furthermore, some options to adapt the program to your specific needs are also available in this setting (see [More Paramters](https://sterniii3.github.io/drugdevelopR/articles/More_Parameters.html)), however skipping phase II, setting predefined benefit categories and choosing different treatment effects in phase II and III are not possible.

* `hr1` is our hazard ratio for the treatment effect of the endpoint OS and `hr2` is our hazard ratio for the treatment effect of the endpoint PFS. We assume that our experimental treatment *exper* leads to unfavorable events occurring only 80\% of times compared to the control treatment *contro* for endpoint OS and only 75\% of times for PFS. Therefore, we set `hr1 = 0.8` and `hr2 = 0.75`.
* `n2min` and `n2max` specify the minimal and maximal number of participants (not events) for the phase II trial. The package will search for the optimal sample size within this region. For now, we want the program to search for the optimal sample size in the interval between 20 and 400 participants. In addition, we will tell the program to search this region in steps of ten participants at a time by setting `stepn2 = 10`.
* Differing from the basis setting, one has to provide to sets of benefits triples, one triple `b11`, `b21`, `b31` in the case endpoint OS is significant (independent of PFS is significant or not) and one triple `b12`, `b22`, `b32` in the case OS is not significant. The intuition is, that overall survival is the "more relevant" endpoint, i.e. if this endpoints shows a significant result, we have higher gains than if only the "less important" endpoint PFS shows a significant result. We assign benefit triples of {1000, 2000, 3000} for the case OS is significant and {1000, 1500, 2000} if not (all in units of 10^5\$).
* Finally, the parameter `rho` determines the correlation between the treatment effects of the two endpoints. In our case, we set `rho = 0.6`, indicating the correlation between the treatment effects of OS and PFS is 0.6.

```{r,eval = FALSE}
res1 <- optimal_multiple_tte(hr1 = 0.8, hr2 = 0.75,                   # define assumed true HRs
                    id1 = NULL, id2 = NULL,
                    n2min = 20, n2max = 400, stepn2 = 10,             # define optimization set for n2
                    hrgomin = 0.7, hrgomax = 0.9, stephrgo = 0.05,    # define optimization set for HRgo
                    alpha = 0.05, beta = 0.1,                         # drug development planning parameters
                    c2 = 0.75, c3 = 1, c02 = 100, c03 = 150,          # define costs for phase II and III
                    K = Inf, N = Inf, S = -Inf,                       # set constraints
                    steps1 = 1,  stepm1 = 0.95,  stepl1 = 0.85,       # effect size categories
                    b11 = 1000, b21 = 2000, b31 = 3000,
                    b12 = 1000, b22 = 1500, b32 = 2000,               # define expected benefits (both categories)
                    rho = 0.6, fixed = TRUE,                          # correlation and treatment effect
                    num_cl = 2)
```

```{r, eval=TRUE, include=FALSE}
# Comment this chunk after running it once
#res1 <- optimal_multiple_tte(hr1 = 0.8, hr2 = 0.75,                   # define assumed true HRs
#                    id1 = NULL, id2 = NULL,
#                    n2min = 20, n2max = 400, stepn2 = 10,             # define optimization set for n2
#                    hrgomin = 0.7, hrgomax = 0.9, stephrgo = 0.05,    # define optimization set for HRgo
#                    alpha = 0.05, beta = 0.1,                         # drug development planning parameters
#                    c2 = 0.75, c3 = 1, c02 = 100, c03 = 150,          # define costs for phase II and III
#                    K = Inf, N = Inf, S = -Inf,                       # set constraints
#                    steps1 = 1,  stepm1 = 0.95,  stepl1 = 0.85,       # effect size categories
#                    b11 = 1000, b21 = 2000, b31 = 3000,
#                    b12 = 1000, b22 = 1500, b32 = 2000,               # define expected benefits (both categories)
#                    rho = 0.6, fixed = TRUE,                          # correlation and treatment effect
#                    num_cl = 2)
#saveRDS(res1, file="optimal_multiple_tte.RDS")
```
```{r, eval=TRUE, include=FALSE}
res1 <- readRDS(file="optimal_multiple_tte.RDS")
```

## Interpreting the output

After setting all these input parameters and running the function, let's take a look at the output of the program.

```{r}
res1
```
The program returns a data frame, with the following results:

* `res$n2` is the optimal number of participants for phase II and `res$n3` the resulting number of events for phase III. We see that the optimal scenario requires 140 participants in phase II and 284 participants in phase III.
* `res$HRgo` is the optimal threshold value for the go/no-go decision rule. We see that we need a hazard ratio of less than 0.85 in phase II in order to proceed to phase III. 
* `res$u` is the expected utility of the program for the optimal sample size and threshold value. In our case it amounts to 295.29, i.e we have an expected utility of 29 529 000\$.
* `res$u` is the probability that endpoint OS is significant, given that the program is successful overall. The probability of a successful program is 0.49, if the program is successful the probability that the endpoint OS is significant and, thus, we obtain higher benefits is 0.77.

# Both endpoints have to be significant



# Where to go from here
In this article we presented examples when multiple endpoints are of interest.
For more information on how to use the package, see:

* [*Introduction to drugdevelopR:*](https://sterniii3.github.io/drugdevelopR/articles/introduction-to-drugdevelopR.html) See how the package works in a basic example.
* *Different outcomes:* Apply it to [binary endpoints](https://sterniii3.github.io/drugdevelopR/articles/Binary_distributed_outcomes.html) and [time-to-event endpoints](https://sterniii3.github.io/drugdevelopR/articles/Time-to-event_outcomes.html).
* [*Interpreting the rest of the output:*](https://sterniii3.github.io/drugdevelopR/articles/Interpreting_Output.html) Obtain further details on your drug development program.
* [*Fixed or prior:*](https://sterniii3.github.io/drugdevelopR/articles/Fixed_and_prior_distributions.html) Model the assumed treatment effect on a prior distribution.
* [*More parameters:*](https://sterniii3.github.io/drugdevelopR/articles/More_Parameters.html) Define custom effect size categories. Put constraints on the optimization by defining maximum costs, the total expected sample size of the program or the minimum expected probability of a successful program. Define an expected difference in treatment effect between phase II and III. Skip phase II.
* *Complex drug development programs:* Adapt to situations with [biased effect estimators](https://sterniii3.github.io/drugdevelopR/articles/Bias_adjustment.html), [multiple phase III trials](https://sterniii3.github.io/drugdevelopR/articles/Multitrial.html) or [multi-arm trials](https://sterniii3.github.io/drugdevelopR/articles/Multiarm_Trials.html).
* [*Parallel computing:*](https://sterniii3.github.io/drugdevelopR/articles/More_Parameters.html) Be faster at calculating the optimum by using parallel computing.
