---
title: "Multiple_Endpoints"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multiple_Endpoints}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(drugdevelopR)
```

For more information on how to use the package, see:

* [*Introduction to drugdevelopR:*](https://sterniii3.github.io/drugdevelopR/articles/introduction-to-drugdevelopR.html) See how the package works in a basic example.
* *Different outcomes:* Apply it to binary endpoints and time-to-event endpoints.
* [*Interpreting the rest of the output:*](https://sterniii3.github.io/drugdevelopR/articles/Interpreting_Output.html) Obtain further details on your drug development program.
* [*Fixed or prior:*](https://sterniii3.github.io/drugdevelopR/articles/Fixed_and_prior_distributions.html) Model the assumed treatment effect on a prior distribution.
* [*More parameters:*](https://sterniii3.github.io/drugdevelopR/articles/More_Parameters.html) Define custom effect size categories. Put constraints on the optimization by defining maximum costs, the total expected sample size of the program or the minimum expected probability of a successful program. Define an expected difference in treatment effect between phase II and III. Skip phase II.
* *Complex drug development programs:* Adapt to situations with biased effect estimators, multiple phase III trials, multi-arm trials, or multiple endpoints.
* *Parallel computing:* Be faster at calculating the optimum by using parallel computing.
